---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/16/22"
author: "Narissa Jimenez-Petchumrus"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\



```{r}
#Load/install required package here
library(tseries)
library(forecast)
library(knitr)
library(sarima)

```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 

\begin{enumerate}[label=(\alph*)]

\item AR(2)

> Answer: For an AR model overall, the series current values depend on its own previous values. The ACF of an AR model shows the lags decreasing gradually over time and also if the first lag displays positive autocorrelation, then an AR model fits best. The PACF of the AR model assists us in identifying the order of the AR model. What this resembles is the first lag spikes downward/negative autocorrelation, the PACF "cuts off" at the order of the AR model. In this case, at lag 2 is where the lag is cut off.

\item MA(1)

> Answer: For an MA model overall, the current deviation from the mean depends on previous deviations from the mean. The ACF of an MA model assists us in identifying the order of the MA model. What this resembles if the first lag spikes downward/negative autocorrelation, the ACF "cuts off" at the order of the model. In this case, at lag 1 is where the lag is cut off.

\end{enumerate}

## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r create ARMA}
ARMA10<-arima.sim(model=list(ar=0.6),n=100)
ARMA10

ARMA01<-arima.sim(model=list(ma=0.9),n=100)
ARMA01

ARMA11<-arima.sim(model=list(ar=0.6,ma=0.9),n=100)
ARMA11


```


\begin{enumerate}[label=(\alph*)]

\item Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r plot ACF}
par(mfrow=c(1,3))
Acf(ARMA10,lag.max=40)
Acf(ARMA01,lag.max=40)
Acf(ARMA11,lag.max=40)

```


\item Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r plot PACF}
par(mfrow=c(1,3))
Pacf(ARMA10,lag.max=40)
Pacf(ARMA01,lag.max=40)
Pacf(ARMA11,lag.max=40)

```

\item Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: By looking at the ACFs and PACFs, it would be feasible to identify the models and orders of each component. We would be able to identify an AR model such as ARMA(1,0) by seeing its ACF lags gradually decrease and its PACF determining the model order since the lags drop off after lag 1, thus the order is p=1. We would be able to identify the MA model such as ARMA(0,1) by seeing its ACF determine the model order q=1 since the lags drop off immediately after lag 1. The PACF reveals the lags exponentially decaying. The ARMA (1,1) is a little harder to identify just by looking at its ACF/PACF as ARMA embraces qualities of both an AR and MA model. For example with the ARMA(1,1) its PACF lags exponentially decay like an MA model and also its ACF shares traits of both ARMA(1,0) and ARMA(0,1) as it somewhat has some of the gradual lag decrease of ARMA(1,0), but also has a lag cutoff appearance similar to ARMA(1,0). It's not super clearcut by just looking at ACF and PACF for ARMA(1,1) 

\item Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: The PACF plot matches for ARMA(1,0) as the lag determining the model's order (p=1) for the PACF is appears to be at the coefficient value 0.6.


\item Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r create second ARMA}
ARMA102<-arima.sim(model=list(ar=0.6),n=1000)
ARMA102

ARMA012<-arima.sim(model=list(ma=0.9),n=1000)
ARMA012

ARMA112<-arima.sim(model=list(ar=0.6,ma=0.9),n=1000)
ARMA112


```

```{r plot second ARMA ACF}
par(mfrow=c(1,3))
Acf(ARMA102,lag.max=40)
Acf(ARMA012,lag.max=40)
Acf(ARMA112,lag.max=40)

```

```{r plot second ARMA PACF}
par(mfrow=c(1,3))
Pacf(ARMA102,lag.max=40)
Pacf(ARMA012,lag.max=40)
Pacf(ARMA112,lag.max=40)
```

\end{enumerate}

> Answer: By looking at the ACFs and PACFs while adding more observations (n=1000), it made it even more feasible to identify the models, orders of each component, and phi/theta. We would be able to identify an AR model such as ARMA(1,0) by seeing its ACF lags gradually decrease and its PACF determining the model order since the lags drop even more dramatically off after lag 1, thus the order is p=1. We would be able to identify the MA model such as ARMA(0,1) by seeing its ACF determine the model order q=1 since the lags drop off even more immediately after lag 1. The PACF reveals the lags exponentially decaying. The ARMA (1,1) seems to be even harder to identify with more observations just by looking at its ACF/PACF as ARMA embraces qualities of both an AR and MA model.It's PACF experiences exponential decay like an MA model and its ACF gradually decreases like an AR model. 

> Answer: The PACF plot matches for ARMA(1,0) even more closely with n=1000 as the lag determining the model's order (p=1) for the PACF appears to be exactly at the coefficient value 0.6.

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

\begin{enumerate}[label=(\alph*)]

\item Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

\item Also from the equation  what are the values of the parameters, i.e., model coefficients. 

\end{enumerate}

> Answer: SARIMA(p=1,d=0,q=1)(P=1,D=0,Q=0)[s=12], model coefficients phi= 0.7, theta= 0.1, SAR coefficient is -0.25.

## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r create sarima model}
sarima_model<-sim_sarima(model=list(ma=0.5,sar=0.8,nseasons=12),n=1000)
sarima_model

```

```{r plot SARIMA ACF}
par(mfrow=c(1,2))
acf(sarima_model,lag.max=40)
pacf(sarima_model,lag.max=40)

```

> Answer: The plots represent well the model we simulated. There are positive spikes in ACF at lag 12, 24, 36 and a single positive spike in PACF at lag 12, which the order of the seasonal component is then P=1 and that makes Q=0 since (P+Q) cannot be greater than 1. This also means that this is a SAR process. Looking at the ACF of the non-seasonal lags, there doesn't appear to be long memory present or a slow decrease of lags versus a cutoff right after lag 1. Therefore it's an MA process where p=1 and q=0.
