---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 3 - Due date 02/08/22"
author: "Narissa Jimenez-Petchumrus"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r load packages, include=FALSE}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(lubridate)
library(stringr)
library(ggplot2)

```

```{r clear global environment, include=FALSE}
rm(list=ls()) ##This code here removes all the objects from your Global Environment.

```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r load data}
energy_data <- read_xlsx(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 
energy_data

colnames(energy_data) <- read_col_names
head(energy_data)

nobs <- nrow(energy_data) 

sapply(energy_data, class)

#create data frame with date and 3 energy types
energy_data2 <- energy_data[, c(1,4:6)] %>% as.data.frame()
energy_data2
sapply(energy_data2, class)

#change column names
names(energy_data2) <- c('Month','Total_Biomass_Energy_Production','Total_Renewable_Energy_Production','Hydroelectric_Power_Consumption')
head(energy_data2)

#Create time series object
ts_energy <- ts(energy_data[4:6]) # TODO: add stuff
ts_energy %>% head()


```

```{r ACF PACF}
#Plot ACF and PACF
for(i in 1:ts_energy){
  par(mfrow=c(1,2))  #place plot side by side
  Acf(ts_energy[,i],lag.max=40,main=paste("ACF of EIA Data",i,sep="")) 
  Pacf(ts_energy[,i],lag.max=40,main=paste("PACF of EIA Data",i,sep=""))
}

```

```{r plots}
#using package ggplot2

ggplot(energy_data2, aes(x=Month, y=Trillion_Btus)) +
            geom_line(aes(y=Total_Renewable_Energy_Production,col="Total_Renewable_Energy_Production")) +
            geom_line(aes(y=Hydroelectric_Power_Consumption,col="Hydroelectric_Power_Consumption"))+
            geom_line(aes(y=Total_Biomass_Energy_Production,col="Total_Biomass_Energy_Production")) + 
            xlab("Time") +
            labs(title = "EIA Dec 2021 Energy Review", color="Energy Types")


```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?



### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.


```{r Biomass regression}

#Create vector t
t <- c(1:nobs)

#Fit a linear trend to TS
lm_model=lm(Total_Biomass_Energy_Production~t,data=energy_data2)
summary(lm_model)
beta0=as.numeric(lm_model$coefficients[1])  #first coefficient is the intercept term or beta0
beta1=as.numeric(lm_model$coefficients[2])  #second coefficient is the slope or beta1

#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Total_Biomass_Energy_Production)) +
            geom_line(color="blue") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
   geom_line(aes(y=detrend_inflow_data), col="green")+
            ggtitle("Total Biomass Energy Production") +
            xlab("Time")

```

```{r renewable regression}
#Fit a linear trend to TS
lm_model2=lm(Total_Renewable_Energy_Production~t,data=energy_data2)
summary(lm_model2)
beta02=as.numeric(lm_model2$coefficients[1])  #first coefficient is the intercept term or beta0
beta12=as.numeric(lm_model2$coefficients[2])  #second coefficient is the slope or beta1


#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Total_Renewable_Energy_Production)) +
            geom_line(color="green") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta02, slope = beta12, color="red") +
            geom_smooth(color="red",method="lm") +
            ggtitle("Total Renewable Energy Production") +
            xlab("Time")

```

```{r hydro regression}
#Fit a linear trend to TS
lm_model3=lm(Hydroelectric_Power_Consumption~t,data=energy_data2)
summary(lm_model3)
beta03=as.numeric(lm_model3$coefficients[1])  #first coefficient is the intercept term or beta0
beta13=as.numeric(lm_model3$coefficients[2])  #second coefficient is the slope or beta1


#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Hydroelectric_Power_Consumption)) +
            geom_line(color="black") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta03, slope = beta13, color="red") +
            geom_smooth(color="red",method="lm") +
            ggtitle("Total Hydroelectric Power Consumption") +
            xlab("Time")

```

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r detrend biomass}
#remove the trend from series
detrend_biomass_data <- energy_data2[,2]-(beta0+beta1*t)
     
#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Total_Biomass_Energy_Production)) +
            geom_line(color="blue") +
            ylab("Trillion Btus") +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_biomass_data), col="green") +
            geom_smooth(aes(y=detrend_biomass_data),color="orange",method="lm")

            
```

```{r detrend renewables}
#remove the trend from series
detrend_renewables_data <- energy_data2[,3]-(beta02+beta12*t)

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Total_Renewable_Energy_Production)) +
            geom_line(color="blue") +
            ylab(paste0("Trillion Btus")) +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_renewables_data), col="green") +
            geom_smooth(aes(y=detrend_renewables_data),color="orange",method="lm") 


```

```{r detrend hydro}
#remove the trend from series
detrend_hydro_data <- energy_data2[,4]-(beta03+beta13*t)

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=energy_data2[,4])) +
            geom_line(color="blue") +
            ylab("Trillion Btus") +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_hydro_data), col="green") +
            geom_smooth(aes(y=detrend_hydro_data),color="orange",method="lm") 


```

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

```{r}


```

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?


