---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 3 - Due date 02/08/22"
author: "Narissa Jimenez-Petchumrus"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r load packages, include=FALSE}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(readxl)
library(lubridate)
library(stringr)
library(ggplot2)

```


##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r load data, include=FALSE}
energy_data <- read_xlsx(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 
energy_data
```

```{r create data frame, include=FALSE}

colnames(energy_data)=c("Month","Total_Biomass_Energy_Production","Total_Renewable_Energy_Production", "Hydroelectric_Power_Consumption")
head(energy_data)

nobs <- nrow(energy_data) 



sapply(energy_data, class)

```

```{r create time series object, include=FALSE}
#create data frame with date and 3 energy types
energy_data2 <- energy_data[, c(1,4:6)] %>% as.data.frame()
energy_data2
sapply(energy_data2, class)

#change column names
names(energy_data2) <- c('Month','Total_Biomass_Energy_Production','Total_Renewable_Energy_Production','Hydroelectric_Power_Consumption')
head(energy_data2)

#Create time series object
ts_energy <- ts(energy_data[4:6], frequency=12)
ts_energy %>% head()

ntype<-ncol(energy_data2)-1

```

```{r acf pacf, echo=FALSE, warning=FALSE, message=FALSE}
#Plot ACF and PACF

for(i in 1:ntype){
  par(mfrow=c(1,3))  #place plot side by side
  
  plot(ts_energy[,i], ylab="Generation in Trillion Btu", main=paste0(colnames(energy_data)[(1+i)]))
  abline(h=mean(ts_energy[,i]),col="red")    
  
  Acf(ts_energy[,i],lag.max=40,main=paste0(colnames(energy_data)[(1+i)])) 
  Pacf(ts_energy[,i],lag.max=40,main=paste0(colnames(energy_data)[(1+i)]))
  
  
}

```

```{r plots, echo=FALSE, warning=FALSE, message=FALSE}
#using package ggplot2

ggplot(energy_data2, aes(x=Month, y=Trillion_Btus)) +
            geom_line(aes(y=Total_Renewable_Energy_Production,col="Total_Renewable_Energy_Production")) +
            geom_line(aes(y=Hydroelectric_Power_Consumption,col="Hydroelectric_Power_Consumption"))+
            geom_line(aes(y=Total_Biomass_Energy_Production,col="Total_Biomass_Energy_Production")) + 
            xlab("Time") +
            labs(title = "EIA Dec 2021 Energy Review", color="Energy Types")


```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?


From the plots in Q1, it appears that both biomass and renewable energy production both are upward trending over time. Hydroelectric power consumption doesn't appear to have a trend. 


### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.


For Biomass, the intercept = 134.78 and the slope = 0.474. This shows that approximately for every month, there is an increase of 0.474 trillion Btus in Biomass energy generation. 

For Hydroelectric, the intercept = 259.18 and the slope = -0.079. This shows that approximately for every month, there is a decrease of 0.079 trillion Btus in Hydro energy consumption. 

For Renewables, the intercept = 323.18 and the slope = 0.88. This shows that approximately for every month, there is a decrease of 0.88 trillion Btus in Renewable energy generation. 


```{r Biomass regression, echo=FALSE, warning=FALSE, message=FALSE}

#Create vector t
t <- c(1:nobs)

#Fit a linear trend to TS
lm_model=lm(Total_Biomass_Energy_Production~t,data=energy_data2)
summary(lm_model)
beta0=as.numeric(lm_model$coefficients[1])  #first coefficient is the intercept term or beta0
beta1=as.numeric(lm_model$coefficients[2])  #second coefficient is the slope or beta1

#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Total_Biomass_Energy_Production)) +
            geom_line(color="blue") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            ggtitle("Total Biomass Energy Production") +
            xlab("Time")

```

```{r renewable regression, echo=FALSE, warning=FALSE, message=FALSE}
#Fit a linear trend to TS
lm_model2=lm(Total_Renewable_Energy_Production~t,data=energy_data2)
summary(lm_model2)
beta02=as.numeric(lm_model2$coefficients[1])  #first coefficient is the intercept term or beta0
beta12=as.numeric(lm_model2$coefficients[2])  #second coefficient is the slope or beta1


#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Total_Renewable_Energy_Production)) +
            geom_line(color="green") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta02, slope = beta12, color="red") +
            geom_smooth(color="red",method="lm") +
            ggtitle("Total Renewable Energy Production") +
            xlab("Time")

```

```{r hydro regression, echo=FALSE, warning=FALSE, message=FALSE}
#Fit a linear trend to TS
lm_model3=lm(Hydroelectric_Power_Consumption~t,data=energy_data2)
summary(lm_model3)
beta03=as.numeric(lm_model3$coefficients[1])  #first coefficient is the intercept term or beta0
beta13=as.numeric(lm_model3$coefficients[2])  #second coefficient is the slope or beta1


#Plot the time series with its trend line
ggplot(energy_data2, aes(x=Month, y=Hydroelectric_Power_Consumption)) +
            geom_line(color="black") +
            ylab(paste0("Trillion Btus")) +
            #geom_abline(intercept = beta03, slope = beta13, color="red") +
            geom_smooth(color="red",method="lm") +
            ggtitle("Total Hydroelectric Power Consumption") +
            xlab("Time")

```

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?


Regarding biomass, the detrended series isn't showing that dramatic of an upward trend over time. In addition the mean of the detrended series is approximately at y=0. It also appears easier to see seasonality if there is any.

Regarding renewables, the detrended series is also not showing that dramatic of an upward trend over time. The mean of this detrended series is also at approximately y=0. It also appears easier to see seasonality if there is any.

Regarding hydroelectric, the detrended series went from a slightly negative trend (based on just staring at the graph though) to showing no potential trend (thus detrended but again this is by only looking at the graph). The mean of this detrended series is also approximately at y=0. It also appears easier to see seasonality if there is any.


```{r detrend biomass, echo=FALSE, warning=FALSE, message=FALSE}
#remove the trend from series
detrend_biomass_data <- energy_data2[,2]-(beta0+beta1*t)
     
#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Total_Biomass_Energy_Production)) +
            geom_line(color="blue") +
            ylab("Trillion Btus") +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_biomass_data), col="green") +
            geom_smooth(aes(y=detrend_biomass_data),color="orange",method="lm")

            
```

```{r detrend renewables, echo=FALSE, warning=FALSE, message=FALSE}
#remove the trend from series
detrend_renewables_data <- energy_data2[,3]-(beta02+beta12*t)

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Total_Renewable_Energy_Production)) +
            geom_line(color="blue") +
            ylab(paste0("Trillion Btus")) +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_renewables_data), col="green") +
            geom_smooth(aes(y=detrend_renewables_data),color="orange",method="lm") 


```

```{r detrend hydro, echo=FALSE, warning=FALSE, message=FALSE}
#remove the trend from series
detrend_hydro_data <- energy_data2[,4]-(beta03+beta13*t)

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=energy_data2[,4])) +
            geom_line(color="blue") +
            ylab("Trillion Btus") +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_hydro_data), col="green") +
            geom_smooth(aes(y=detrend_hydro_data),color="orange",method="lm") 


```

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?


Looking at the plots comparing PACF and ACF of the trend and detrended series, there isn't that much of a dramatic change between the ACFs and PACFs. The only ones that appear to be different between the trend and detrended series are renewables ACF (the detrended renewables ACF shows a little seasonality compared to trend renewable ACF), biomass ACF (the detrended ACF shos a little less of a gradual decline compared to the trend ACF, potentially some seasonality), biomass PACF (lag 12 on the detrended series looks barely a little higher unless my eyesight sucks), and renewables PACF (a few of the lags on the detrended PACF looks taller compared to the trend PACF which again unless my eyesight sucks).



```{r detrend ACF PACF, echo=FALSE, warning=FALSE, message=FALSE}
#create time series of detrended data
ts_detrend_biomass<- ts(detrend_biomass_data, start = 1973, frequency = 12)
ts_detrend_renewable<- ts(detrend_renewables_data, start = 1973, frequency = 12)
ts_detrend_hydro<- ts(detrend_hydro_data, start = 1973, frequency = 12)

#Comparing ACF and PACF of each series before and after detrending
    
par(mfrow=c(3,2)) #Comparing ACFs
   
  Acf(ts_detrend_biomass,lag.max=40,main="Biomass Detrend ACF") 
  Acf(ts_energy[,1],lag.max=40,main="Biomass ACF")
  
  Acf(ts_detrend_renewable,lag.max=40,main="Renewable Detrend ACF") 
  Acf(ts_energy[,2],lag.max=40,main="Renewable ACF")
  
  Acf(ts_detrend_hydro,lag.max=40,main="Hydro Detrend ACF") 
  Acf(ts_energy[,3],lag.max=40,main="Hydro ACF")
  
#Comparing PACFs
  
par(mfrow=c(3,2)) #Comparing PACFs
   
  Pacf(ts_detrend_biomass,lag.max=40,main="Biomass Detrend PACF") 
  Pacf(ts_energy[,1],lag.max=40,main="Biomass PACF")
  
  Pacf(ts_detrend_renewable,lag.max=40,main="Renewable Detrend PACF") 
  Pacf(ts_energy[,2],lag.max=40,main="Renewable PACF")
  
  Pacf(ts_detrend_hydro,lag.max=40,main="Hydro Detrend PACF") 
  Pacf(ts_energy[,3],lag.max=40,main="Hydro PACF")


```


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.



The series that has an obvious seasonal trend is hydroelectric based on plots and graphs. During mostly non-summer months, there appears to be more hydroelectric resources available as the regression coefficients are higher as there's more precipitation around this time. Late summer to fall there appears to be less hydroelectric sources available given these months are dryer. 


```{r seasonal, echo=FALSE, warning=FALSE}
#create the seasonal dummies
dummies <- seasonaldummy(ts_energy[,3])  

#Then fit a linear model to the seasonal dummies
seas_means_model=lm(energy_data2[,4]~dummies)
summary(seas_means_model)

#Store regression coefficients
beta_int=seas_means_model$coefficients[1]
beta_coeff=seas_means_model$coefficients[2:12]

```


### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?


Comparing the seasonal and deseasoned data of hydroelectric power, the deseasoned data shows some values that could potentially be outliers that the seasonal data doesn't reveal. In addition, the mean appears to be at y=0. 


```{r deseasoned, echo=FALSE, warning=FALSE}
#compute seasonal component
energy_seas_comp=array(0,nobs)
for(i in 1:nobs){
  energy_seas_comp[i]=(beta_int+beta_coeff%*%dummies[i,])
}

#Plotting seasonality
ggplot(energy_data2, aes(x=Month, y=Hydroelectric_Power_Consumption)) +
            geom_line(color="blue") +
            ylab("Hydropower Consumption in Trillion Btu") +
            ggtitle("Seasonality in Hydropower generation") +
            geom_line(aes(y=energy_seas_comp), col="red")


#Removing seasonal component
deseason_energy_data <- energy_data2[,(4)]-energy_seas_comp

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Hydroelectric_Power_Consumption)) +
            geom_line(color="blue") +
            ylab("Hydropower Consumption in Trillion Btu") +
            ggtitle("Hydropower Consumption with and without seasonality") +
            geom_line(aes(y=deseason_energy_data), col="green")


```

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?


Seasonality for hydroelectric appears to be much more controlled for comparing the deaseasoned ACF with the seasonal ACF. Regarding the PACFs, the deseasoned hydroelectric graph shows shorter lags compared to the seasonal PACF.


```{r deseasoned ACF PACF, echo=FALSE, warning=FALSE}
ts_deseason_energy<- ts((deseason_energy_data), start = 1973, frequency = 12)


par(mfrow=c(2,2))
Acf(ts_deseason_energy,lag.max=40,main="Deseasoned Hydro ACF")
Acf(ts_energy[,3],lag.max=40,main="Hydro ACF with seasonality")

pacf(ts_deseason_energy,lag.max=40,main="Deseasoned Hydro PACF")
Pacf(ts_energy[,3],lag.max=40,main="Hydro PACF with seasonality")



```
