---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 4 - Due date 02/17/22"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp21.Rmd"). Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r load packages, include=FALSE}
#Load/install required package here
library(readxl)
library(ggplot2)
library(forecast)
library(tseries)
library(Kendall)
library(base)
library(dplyr)
library(tidyr)

```

## Questions 

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production". 


```{r import data, include=FALSE}
#Importing data set - using xlsx package
energy_data <- read_xlsx(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 
energy_data

```

```{r create data frame, include=FALSE}

colnames(energy_data)=c("Month","Total_Biomass_Energy_Production","Total_Renewable_Energy_Production", "Hydroelectric_Power_Consumption")
head(energy_data)

nobs <- nrow(energy_data) 
sapply(energy_data, class)

#create data frame with date and 3 energy types
energy_data2 <- energy_data[, c(1,4:6)] %>% as.data.frame()
energy_data2
sapply(energy_data2, class)

#change column names
names(energy_data2) <- c('Month','Total_Biomass_Energy_Production','Total_Renewable_Energy_Production','Hydroelectric_Power_Consumption')
head(energy_data2)
sapply(energy_data2,class)

```


## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments:
* *x* vector containing values to be differenced;
* *lag* integer indicating with lag to use;
* *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series Do the series still seem to have trend?


The plot doesn't appear to still have a trend after differencing. 

```{r differencing, echo=FALSE}
vector_renewables<-energy_data2$Total_Renewable_Energy_Production
vector_renewables_diff<-diff(vector_renewables, lag=1, differences=1)


ts_renewables<-ts(vector_renewables_diff,start=c(1973,1),end=c(2021,9),frequency=12)
plot(ts_renewables)

```

### Q2

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in A3 using linear regression. (Hint: Just copy and paste part of your code for A3)

Copy and paste part of your code for A3 where you compute regression for Total Energy Production and the detrended Total Energy Production


Comparing the plots from Q1 to Q2, the detrended series in green compared to the differenced series from Q1 still displays more of a trend. This could be because using a linear regression isn't the best way to detrend this series as likely the trend isn't deterministic versus stochastic. If the series is stochastic, difference-stationarity is obtained by differencing the series. This was done in the plot in Q1, which removed more trend than the plot in Q2. 


```{r detrend renewables, echo=FALSE, warning=FALSE, message=FALSE}
#Create vector t
t <- c(1:nobs)

#Fit a linear trend to TS
lm_model2=lm(Total_Renewable_Energy_Production~t,data=energy_data2)
summary(lm_model2)
beta02=as.numeric(lm_model2$coefficients[1])  #first coefficient is the intercept term or beta0
beta12=as.numeric(lm_model2$coefficients[2])  #second coefficient is the slope or beta1

#remove the trend from series
detrend_renewables_data <- energy_data2[,3]-(beta02+beta12*t)

#Understanding what we did
ggplot(energy_data2, aes(x=Month, y=Total_Renewable_Energy_Production)) +
            geom_line(color="blue") +
            ylab(paste0("Trillion Btus")) +
            geom_abline(intercept = beta0, slope = beta1, color="red") +
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_renewables_data), col="green") +
            geom_smooth(aes(y=detrend_renewables_data),color="orange",method="lm") 


```

### Q3
Create a data frame with 4 columns: month, original series, detrended by Regression Series and differenced series. Make sure you properly name all columns. Also note that the differenced series will have only 584 rows because you loose the first observation when differencing. Therefore, you need to remove the first observations for the original series and the detrended by regression series to build the new data frame.

```{r create dataframe, include=FALSE}
#Data frame - remember to not include January 1973

#month + original data minus row 1
orig_renewables_df<-energy_data2[-1,c(1,3)] %>% as.data.frame()
orig_renewables_df

#detrended data minus row 1
detrend_df<-as.data.frame(detrend_renewables_data)
detrend_df
detrend_df=detrend_df[-1,]
detrend_df

#differenced data
vector_diff_df<-as.data.frame(vector_renewables_diff)
vector_diff_df

#smack it all together
all_data_df<-data.frame(orig_renewables_df, detrend_df, vector_diff_df)
all_data_df

#rename columns
names(all_data_df)<-c('Month','Original_Data','Detrended_Data','Differenced_Data')
all_data_df


```

### Q4
Using ggplot() create a line plot that shows the three series together. Make sure you add a legend to the plot.

```{r ggplot, echo=FALSE}
#Use ggplot

ggplot(all_data_df, aes(x=Month, y=all_data_df[,2])) +
            geom_line(aes(color="steelblue2")) +
            geom_line(aes(y=all_data_df[,3], color="turquoise1")) +
            geom_line(aes(y=all_data_df[,4], color="seagreen2")) +
            scale_color_identity(name="Data Type",
                                 breaks=c("steelblue2","turquoise1","seagreen2"),
                                 labels=c("Original Data","Detrended Data","Differenced Data"),
                                 guide = "legend") +
            labs(title="Renewable Energy Series",
                 x="Month",
                 y="Trillions of Btus",
                 ) 

            

```


### Q5
Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the Acf() function to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?


Looking at the ACFs, the differencing was the best at eliminating the trend as both the original and detrended ACFs we can see the lags decrease while the differenced ACF . The detrended ACF also displays seasonality.


```{r ACFs, echo=FALSE}
ts_detrend<-ts(detrend_renewables_data,start=c(1973,1),end=c(2021,9),frequency=12)
ts_original<-ts(energy_data2$Total_Renewable_Energy_Production,start=c(1973,1),end=c(2021,9),frequency=12)

par(mfrow=c(2,2))
Acf(ts_original,lag.max = 40, ylim=c(-0.5,1),main="ACF Original")
Acf(ts_detrend,lag.max = 40, ylim=c(-0.5,1),main="ACF Detrended")
Acf(ts_renewables,lag.max=40, ylim=c(-0.5,1),main="ACF Differenced")


```


### Q6
Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. Whats the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.



According to the Seasonal Mann-Kendall test, the p-value being below my self-declared critical value of 0.05 means that we can reject the null hypothesis (the data is stationary) and embrace the alternative hypothesis. This means that there is a deterministic trend within the data.

According to the ADF Test, I cannot reject the null hypothesis that my series is non-stationary since my p-value is above the critical value of 0.05 (it's 0.8161). This implies that there's a unit root, thus the series does have a stochastic trend. This does match Q2's plot of the regular data series (the blue line that hasn't been detrended).


```{r tests, echo=FALSE}
SeasonalMannKendall(ts_original)
adf.test(ts_original)

```

### Q7

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. 

```{r}

```

### Q8

Apply the Mann Kendall, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the non-aggregated series, i.e., results for Q6?

```{r}

```

